---
title: "MasonHolcombe_408_FinalProject"
format: html
editor: visual
---

Mason Holcombe

Stat 408 Final Project R code for data processing, model building, and visualizations

## Dependencies

```{r echo=FALSE}
library(dplyr)
library(ggplot2)
library(tidyr)
library(glmnet)
library(knitr)
```

```{r}
set.seed(123)
```

## Data Ingestion

```{r}
lol25 = read.csv("Data/2025_LoL_esports_match_data_from_OraclesElixir.csv") %>%
  filter(!is.na(result))
colnames(lol25)
```

## Data Processing 

```{r}
cols_to_remove = c("gameid", "datacompleteness", "url", "league", "year",
                   "split", "date", "game", "participantid",
                   "playername", "playerid", "teamname", "teamid",
                   "ban1", "ban2", "ban3", "ban4", "ban5",
                   "pick1", "pick2", "pick3", "pick4", "pick5",
                   "result")
X = lol25 %>% 
  select(-all_of(cols_to_remove)) %>%
  mutate(across(everything(), ~ replace_na(., 0)))
X = model.matrix(~ . - 1, data = X)
X_df = as.data.frame(X)

y = as.numeric(lol25$result)
```

## LASSO CV

```{r}
lasso_cv = cv.glmnet(
  X, y,
  alpha = 1,
  family = "binomial",
  nlambda = 30,
  nfolds = 5,
)
plot(lasso_cv)
```

#### Optimal Lambda

```{r}
best_lambda = lasso_cv$lambda.min
model_best = glmnet(X, y, alpha = 1, lambda = best_lambda, family = "binomial")
```

#### All Coefficients

```{r}
coef_df = as.data.frame(as.matrix(coef(model_best)))
coef_df$feature = rownames(coef_df)
colnames(coef_df)[1] = "coefficient"

coef_df = coef_df[coef_df$feature != "(Intercept)", ]

ggplot(coef_df, aes(x = reorder(feature, coefficient), y = coefficient)) +
  geom_bar(stat = "identity", fill="steelblue") +
  coord_flip() +
  labs(title = "Coefficients of Lasso Model",
       x = "Features",
       y = "Coefficient Value") +
  theme_minimal()

```

#### Top/Bot 10 Coefficients

```{r}
# Select top 10 positive and top 10 negative
top_coefs = coef_df %>%
  arrange(desc(coefficient)) %>%
  slice(1:20) %>%
  bind_rows(
    coef_df %>%
      arrange(coefficient) %>%
      slice(1:20)
  )

ggplot(top_coefs, aes(x = reorder(feature, coefficient), y = coefficient)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  labs(title = "Top 10 Positive and Negative Coefficients",
       x = "Features",
       y = "Coefficient Value") +
  theme_minimal()

```

```{r}
top_important_lasso_features = top_coefs$feature
top_important_lasso_features
```

## Selected Features Logistic Regression

Model 1: Hand chosen features based empirically\
Model 2: Features determined by LASSO regression

#### Model 1

```{r}
model1_cols = c(
  "playoffs","side","position",
  "kills", "deaths", "assists",
  "teamkills", "teamdeaths",
  "doublekills", "triplekills", "quadrakills", "pentakills",
  "firstblood", "firstbloodkill", "firstbloodassist",
  "team.kpm",
  "damagetochampions", "dpm", "damageshare",
  "damagetakenperminute", "damagemitigatedperminute", "damagetotowers",
  "wardsplaced", "wpm", "wardskilled", "wcpm", "controlwardsbought",
  "visionscore", "vspm",
  "totalgold", "earnedgold", "earned.gpm", "earnedgoldshare", "goldspent",
  "minionkills", "monsterkills",
  "result"
)

model1_data = lol25 %>%
  select(all_of(model1_cols)) %>%
  mutate(across(everything(), ~ replace_na(., 0)))

model1_matrix = model.matrix(~ . - 1, data = model1_data)  # remove intercept
model1_df = as.data.frame(model1_matrix)

model1 = glm(result ~ ., data = model1_df, family = binomial())
```

Print confusion matrix of the binary regression

```{r}
pred_prob1 = predict(model1, type = "response")
pred_class1 = ifelse(pred_prob1 > 0.5, 1, 0)

confusion_matrix1 = as.data.frame.matrix(
  table(pred = pred_class1, actual = model1_df$result)
)

confusion_matrix1
```

#### Model 2

```{r}
model2_cols = c(top_coefs$feature, "result")

model2_data = lol25 %>%
  mutate(across(everything(), ~ replace_na(., 0)))

model2_matrix = model.matrix(~ . - 1, data = model2_data)
model2_matrix = model2_matrix[, model2_cols, drop = FALSE]
model2_df = as.data.frame(model2_matrix)

model2 = glm(result ~ ., data = model2_df, family = binomial())
```

Print confusion matrix of the binary regression

```{r}
pred_prob2 = predict(model2, type = "response")
pred_class2 = ifelse(pred_prob2 > 0.5, 1, 0)

confusion_matrix2 = as.data.frame.matrix(
  table(pred = pred_class2, actual = model2_df$result)
)

confusion_matrix2
```

## Best Model Summary

```{r}
summary(model1)
```

## Visualizations

Win/Loss count in dataset

```{r}
ggplot(lol25, aes(x = factor(result))) +
  geom_bar() +
  labs(title = "Wins vs Losses",
       x = "Result (0 = Loss, 1 = Win)",
       y = "Count")

```

Kills, Deaths, Assists by Result & Position

```{r}
df_long = lol25 %>%
  filter(position != "team") %>%
  select(result, position, kills, deaths, assists) %>%
  pivot_longer(
    cols = c(kills, deaths, assists),
    names_to = "stat",
    values_to = "value"
  ) %>% 
  mutate(
    stat = factor(stat,
                  levels = c("kills", "deaths", "assists"))
  )

ggplot(df_long, aes(x = position, y = value, fill = stat)) +
  geom_boxplot() +
  facet_wrap(~ result,
             labeller = as_labeller(c(`0` = "Loss",
                                      `1` = "Win"))) +
  labs(
    title = "Kills, Deaths, and Assists by Result & Position",
    x = "Position",
    y = "Value"
  ) +
  scale_fill_manual(values = c(
    "kills" = "skyblue",
    "deaths" = "salmon",
    "assists" = "lightgreen"
  ))
```
